\textbf{Beasts of Prey or Rational Animals?\\ Private Governance in Brazil's \textit{Jogo do Bicho}}

### 2018-11-10 paper-bicho/bicho.tex
more updates

Yet the game has survived. The \textit{jogo do bicho} has outlasted more than 30 Brazilian presidents and thrived under both military regimes and democratic governments \citep{jupiara2015poroes}. But more than a act of civic disobedience, the \textit{jogo do bicho} is a very successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}. A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), what makes it the largest clandestine gambling game in the world \citep{huff2015jogos}. \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost as many as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the total Brazilian workforce \citep{time1966bicho}.}

@@

The \textit{jogo do bicho's} market formation closely resembles what \citet{fligstein1996markets} calls `markets as politics', in which firms create institutions to restrain the competition and organise the labour force. At the formation of the lottery market, competition is fierce and incumbents an challengers are akin to social movements. They are constantly trying to convince others of the viability of their ideas. When markets stabilise, however, incumbents collude to impose their conditions of control to other players and to workers. In the case of the \textit{jogo}, this implies in a mix of financial incentives and violence threats carried by corrupt state agents or private bodyguards.

@@

Legal lotteries employ many techniques to show that this is the case. For instance, the Brazilian official lottery, run by the federal government, are regularly audited by the \textit{Controladoria Geral da União} (Comptroller General of Brazil), the \textit{Tribunal de Contas da União} (General Accounting Office), and by Ernst \& Young. The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute \citep{uol2016loteria}. A clandestine lottery, in contrast, cannot provide the same signals. Thus, the game providers need to assure gamblers that their business is honest although it is illegal. \textit{Bicheiros} addressed this issue using a traditional commercial practice: they invested on their reputations.

@@

The funding of the samba schools had an indirect effect to the animal game. The patronage also reduced agent-principal problems within the \emph{jogo do bicho}. \emph{Bicheiros} donate to samba school to gather support of the communities, and by doing so they gain access to local information on their business. Clients who have a positive image of the \emph{bicheiro} may denounce fraudsters to their superiors, thus monitoring is cost-effective for animal game managers. Thus, street bookmakers have fewer incentives to cheat. In addition, street sellers are often recruited from the poor communities, so they tend to be immediate beneficiaries of \emph{bicheiros}'s donations \citep{bbc2012aniz}. Hence, funds donated to samba schools and other charities organisations help align the interests of different members of the \emph{jogo do bicho} organisation. The patronage can be interpreted as an illegal version of `profit-sharing', a mechanism which has induced effectively cooperative behaviour in both small and large corporations \citep{cahuc1997profit,fitzroy1987cooperation, kruse1992profit}.

@@

The Brazilian political system is particularly conductive to clientelistic practices. Brazil has one of the most fragmented party systems in the world, which induces political entrepreneurs to run highly individualised campaigns \citep{figueiredo2000presidential,geddes1992institutional}. In addition, Brazil uses a open-list proportional representation electoral system, that is, each of the 27 states of the federation are considered as at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp[483]{samuels2000ambition}}. These two elements indicate that Brazilian politicians are often free from the strong requirements of political parties and can run their campaigns with a high degree of independence. Nevertheless, that independence means candidates rely mostly on themselves to raise funds and mobilise potential voters. Hence, political campaigns in Brazil tend to be expensive and personality-centred.

@@

Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population \citep[e.g.][]{gambetta1996sicilian,skarbek2011governance,skarbek2012prison,varese2001russian,varese2011mafias}. Yet, the \textit{jogo do bicho} offers a convincing example that it is possible for an illegal syndicate to operate with low levels of violence for more than a hundred years. \textit{Bicheiros} employ a number of strategies to obtain reliable information from their subordinates while offering club goods and other selected benefits to workers. Furthermore, by investing in the Carnival parade \textit{bicheiros} have been able to gather popular and government support. Poor communities have associated with the \textit{bicheiros} to receive welfare provision, whereas politicians have collaborated with them to reap the financial and electoral benefits the \textit{jogo do bicho}'s networks can provide.

Nevertheless, the \textit{jogo do bicho} has also created negative externalities. Violence is used to punish defectors and to constrain competitors. The clientelistic relationship that \textit{bicheiros} have with local politicians have lead to sub-optimal outcomes, such as predatory political campaigning, distortions in electoral representation, and impunity for human rights violations. These negative externalities have long-term effects and still impact the Brazilian public sphere.

>>>>

Yet the game has survived. The \textit{jogo do bicho} has outlasted more than 30 Brazilian presidents and thrived under both military regimes and democratic governments \citep{jupiara2015poroes}. But more than an act of civil disobedience, the \textit{jogo do bicho} is a very successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}. A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), what makes it the largest clandestine gambling game in the world \citep{huff2015jogos}. \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost as many as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the total Brazilian workforce \citep{time1966bicho}.}

@@

The \textit{jogo do bicho's} market formation closely resembles what \citet{fligstein1996markets} calls `markets as politics', in which firms create institutions to restrain the competition and organise the labour force. At the formation of the lottery market, competition is fierce and participants are akin to social movements. They are constantly trying to convince others of the viability of their ideas. When markets stabilise, however, incumbents collude to impose their conditions of control to other players and to workers. In the case of the \textit{jogo}, this implies in a mix of financial incentives and violence threats carried by corrupt state agents or private bodyguards.

@@

Legal lotteries employ many techniques to show that this is the case. For instance, the Brazilian official lottery, run by the federal government, are regularly audited by the \textit{Controladoria Geral da União} (Comptroller General of Brazil), the \textit{Tribunal de Contas da União} (General Accounting Office), and by Ernst \& Young. The balls are measured and weighted every three months by the National Institute of Metrology, Quality and Technology (Inmetro), the Brazilian equivalent of United Kingdom's National Physical Laboratory or the American National Standards Institute \citep{uol2016loteria}.

A clandestine lottery, in contrast, cannot provide the same signals. Thus, the game providers need to assure gamblers that their business is honest although it is illegal. \textit{Bicheiros} addressed this issue using a traditional commercial practice: they invested on their reputations.

@@

The funding of the samba schools had an indirect effect to the animal game. The patronage also reduced agent-principal problems within the \emph{jogo do bicho}. \emph{Bicheiros} donate to samba school to gather support of the communities, and by doing so they gain access to local information on their business. Clients who have a positive image of the \emph{bicheiro} may denounce fraudsters to their superiors, thus monitoring is cost-effective for animal game managers. Thus, street bookmakers have fewer incentives to cheat. In addition, street sellers are often recruited from the poor communities, so they tend to be immediate beneficiaries of \emph{bicheiros}' donations \citep{bbc2012aniz}. Hence, funds donated to samba schools and other charities organisations help align the interests of different members of the \emph{jogo do bicho} organisation. The patronage can be interpreted as an illegal version of `profit-sharing', a mechanism which has induced effectively cooperative behaviour in both small and large corporations \citep{cahuc1997profit,fitzroy1987cooperation, kruse1992profit}.

@@

The Brazilian political system is particularly conductive to clientelistic practices. Brazil has one of the most fragmented party systems in the world, which induces political entrepreneurs to run highly individualised campaigns \citep{figueiredo2000presidential,geddes1992institutional}. In addition, Brazil uses an open-list proportional representation electoral system, that is, each of the 27 states of the federation are considered as at-large electoral districts \citetext{\citealp{ames1995electoral}; \citealp[483]{samuels2000ambition}}. These two elements indicate that Brazilian politicians are often free from the strong requirements of political parties and can run their campaigns with a high degree of independence. Nevertheless, that independence means candidates rely mostly on themselves to raise funds and mobilise potential voters. Hence, political campaigns in Brazil tend to be expensive and personality-centred.

@@

Past research has shown that criminal organisations face considerable challenges to elicit cooperation from their members and establish close ties with the population. Yet, the \textit{jogo do bicho} offers a convincing example that it is possible for an illegal syndicate to operate with low levels of violence for more than a hundred years. \textit{Bicheiros} employ a number of strategies to obtain reliable information from their subordinates while offering club goods and other selected benefits to workers. Furthermore, by investing in the Carnival parade \textit{bicheiros} have been able to gather popular and government support. Poor communities have associated with the \textit{bicheiros} to receive welfare provision, whereas politicians have collaborated with them to reap the financial and electoral benefits the \textit{jogo do bicho}'s networks can provide.

Nevertheless, the \textit{jogo do bicho} has also created negative externalities. Violence is used to punish defectors and to constrain competitors. The clientelistic relationship that \textit{bicheiros} have with local politicians have lead to undemocratic outcomes, such as predatory political campaigning, distortions in electoral representation, and impunity for human rights violations. These negative externalities have long-term effects and still impact the Brazilian public sphere.
